---
title: "STATS 506 — Homework 1"
author: "Zixuan Wei"
format:
  html:
    embed-resources: true
    code-fold: false
execute:
  echo: true
---

GitHub repository: [IvyWeiii/DS506](https://github.com/IvyWeiii/DS506)

# Problem 1: Exploratory Analysis of Soybean Cultivars

## (a) Import the Data

The dataset is imported as a data frame. The column names in the CSV header are retained.

```{r}
# Import the dataset using the first row as column names.
soybean <- read.csv(
  "data/soybean/data.csv",
  header = TRUE,
  stringsAsFactors = FALSE
)

# Treat season and cultivar as categorical variables.
soybean$Season <- factor(soybean$Season)
soybean$Cultivar <- factor(soybean$Cultivar)

# Inspect the imported data.
head(soybean)
str(soybean)
```

## (b) Verify the Dataset Structure

```{r}
# Count observations, distinct cultivars, and seasons.
data.frame(
  Observations = nrow(soybean),
  Cultivars = length(unique(soybean$Cultivar)),
  Seasons = length(unique(soybean$Season))
)

# Display the number of observations in each season.
season_counts <- as.data.frame(table(soybean$Season))
names(season_counts) <- c("Season", "Observations")
knitr::kable(season_counts)

# Check the expected dataset structure.
stopifnot(
  nrow(soybean) == 320,
  length(unique(soybean$Cultivar)) == 40,
  length(unique(soybean$Season)) == 2
)
```

The dataset contains `r nrow(soybean)` observations,
`r length(unique(soybean$Cultivar))` distinct cultivars, and
`r length(unique(soybean$Season))` seasons.
The table above reports the number of observations from each season.

## (c) Explore Relationships and Summaries

### 1. Overall Correlation

```{r}
# Calculate the Pearson correlation between plant height and grain yield.
overall_cor <- cor(soybean$PH, soybean$GY)
overall_cor
```

The overall Pearson correlation between plant height (PH) and grain yield
(GY) is `r round(overall_cor, 4)`.

### 2. Correlation by Season

```{r}
# Calculate the correlation separately for each season.
season_cor <- c(
  cor(
    soybean$PH[soybean$Season == "1"],
    soybean$GY[soybean$Season == "1"]
  ),
  cor(
    soybean$PH[soybean$Season == "2"],
    soybean$GY[soybean$Season == "2"]
  )
)

cor_table <- data.frame(
  Season = c("1", "2"),
  Correlation = season_cor
)
knitr::kable(cor_table, digits = 4)

higher_season <- cor_table$Season[which.max(cor_table$Correlation)]
```

The correlations are `r round(season_cor[1], 4)` for Season 1 and
`r round(season_cor[2], 4)` for Season 2.
Season `r higher_season` has the higher correlation.

### 3. Grain Yield at the Greatest Plant Height

```{r}
# Retain all observations tied for the greatest plant height.
tallest <- soybean[
  soybean$PH == max(soybean$PH),
  c("Season", "Cultivar", "Repetition", "PH", "GY")
]
knitr::kable(tallest, row.names = FALSE)
```

The greatest plant height is `r max(soybean$PH)`.
The corresponding grain yield value(s) are
`r paste(tallest$GY, collapse = ", ")`.

### 4. Seed Weight Above the Overall Average

```{r}
# Calculate the overall mean and percentage strictly above it.
mean_mhg <- mean(soybean$MHG)
pct_above <- 100 * mean(soybean$MHG > mean_mhg)

data.frame(
  Mean_MHG = mean_mhg,
  Percentage_Above_Mean = pct_above
)
```

The overall average MHG is `r round(mean_mhg, 2)`.
Of all observations, `r round(pct_above, 2)`% have an MHG strictly greater
than this average.

### 5. Cultivar with the Highest Average Grain Yield

```{r}
# Calculate mean grain yield for each cultivar across both seasons.
cultivar_means <- aggregate(
  GY ~ Cultivar,
  data = soybean,
  FUN = mean
)

# Retain all cultivars tied for the highest mean.
best_cultivar <- cultivar_means[
  cultivar_means$GY == max(cultivar_means$GY),
]
knitr::kable(best_cultivar, digits = 2, row.names = FALSE)
```

The cultivar(s) with the highest average grain yield are
`r paste(best_cultivar$Cultivar, collapse = ", ")`,
with an average grain yield of `r round(max(cultivar_means$GY), 2)`.

## (d) Overall and Season-Specific Averages

```{r}
# Select quantitative measurements, excluding experimental identifiers.
quant_vars <- c("PH", "IFP", "NLP", "NGP", "NGL", "NS", "MHG", "GY")

# Calculate overall means.
overall_means <- data.frame(
  Group = "Overall",
  as.data.frame(lapply(soybean[quant_vars], mean)),
  check.names = FALSE
)

# Calculate means within each season.
season_means <- aggregate(
  soybean[quant_vars],
  by = list(Group = soybean$Season),
  FUN = mean
)
season_means$Group <- paste("Season", season_means$Group)

# Combine the overall and season-specific results.
mean_table <- rbind(overall_means, season_means)
knitr::kable(mean_table, digits = 2, row.names = FALSE)
```

The table reports the overall and season-specific means of the eight
quantitative measurements. Season, cultivar, and repetition are
experimental identifiers and are not averaged.

## (e) Compare Grain Yield Between Seasons

Let $\mu_1$ and $\mu_2$ denote the population mean grain yields for
Seasons 1 and 2, respectively. The hypotheses are:

- $H_0: \mu_1 = \mu_2$.
- $H_A: \mu_1 \ne \mu_2$.

I use a two-sided Welch two-sample t-test at the 5% significance level.
This analysis assumes independent observations and does not require
equal population variances.

```{r}
# Compare mean grain yield using Welch's two-sample t-test.
welch_test <- t.test(
  GY ~ Season,
  data = soybean,
  alternative = "two.sided",
  var.equal = FALSE
)
welch_test

# Calculate the observed difference: Season 1 minus Season 2.
mean_difference <- mean(soybean$GY[soybean$Season == "1"]) -
  mean(soybean$GY[soybean$Season == "2"])
```

The estimated mean difference (Season 1 minus Season 2) is
`r round(mean_difference, 2)`.
The Welch test gives a t-statistic of
`r round(unname(welch_test$statistic), 4)` and a p-value of
`r format.pval(welch_test$p.value, digits = 4)`.
The 95% confidence interval for the mean difference is
[`r round(welch_test$conf.int[1], 2)`,
`r round(welch_test$conf.int[2], 2)`].

At the 5% significance level, we
`r if (welch_test$p.value < 0.05) "reject" else "fail to reject"`
the null hypothesis. There is
`r if (welch_test$p.value < 0.05) "sufficient" else "insufficient"`
evidence that mean grain yield differs between the two seasons.
A nonsignificant result does not establish that the population means
are equal.

For comparison, I also perform a pooled two-sample t-test, which
additionally assumes equal population variances.

```{r}
# Repeat the comparison under the equal-variance assumption.
pooled_test <- t.test(
  GY ~ Season,
  data = soybean,
  alternative = "two.sided",
  var.equal = TRUE
)
pooled_test
```

The pooled test gives a p-value of
`r format.pval(pooled_test$p.value, digits = 4)`.
The two tests
`r if ((welch_test$p.value < 0.05) == (pooled_test$p.value < 0.05)) "agree" else "disagree"`
on the decision at the 5% significance level.

Both comparisons assume independent observations. Because the same
cultivars appear in both seasons, a design-aware analysis may need
to account for cultivar and experimental blocks.


# Problem 2: Adult Income Dataset

## (a) Import the Data

I use the training file, `adult.data`, for this analysis.
The separate test file, `adult.test`, is not included.

```{r}
# Import the headerless file and recognize question marks as missing.
adult <- read.csv(
  "data/adult/adult.data",
  header = FALSE,
  strip.white = TRUE,
  na.strings = "?",
  stringsAsFactors = FALSE
)

dim(adult)
```

## (b) Clean and Simplify Variable Names

```{r}
# Assign descriptive names in the documented column order.
names(adult) <- c(
  "age", "workclass", "fnlwgt", "education", "education_num",
  "marital_status", "occupation", "relationship", "race", "sex",
  "capital_gain", "capital_loss", "hours_per_week",
  "native_country", "income"
)

head(adult)
str(adult)
```

The variable names use lowercase letters and underscores consistently.
The imported dataset contains `r nrow(adult)` observations and
`r ncol(adult)` variables. Spaces following delimiters are removed,
and question marks are treated as missing values.

## (c) Examine Missing Values

```{r}
# Count missing observations for every variable before filtering.
missing_counts <- data.frame(
  Variable = names(adult),
  Missing = colSums(is.na(adult))
)
knitr::kable(missing_counts, row.names = FALSE)
```

There are `r sum(is.na(adult$workclass))` missing values in workclass,
`r sum(is.na(adult$occupation))` in occupation, and
`r sum(is.na(adult$native_country))` in native country.
The table reports the missing-value count for every variable.
These counts may overlap because one observation can have multiple
missing values.

## (d) Restrict Weekly Working Hours

```{r}
# Keep individuals working at least 10 hours per week.
adult_hours <- subset(adult, hours_per_week >= 10)

data.frame(
  Stage = c("Before filtering", "At least 10 hours per week"),
  Observations = c(nrow(adult), nrow(adult_hours))
)
```

The sample size decreases from `r nrow(adult)` to
`r nrow(adult_hours)` after restricting the sample to individuals
working at least 10 hours per week.

## (e) Remove Missing Values in Selected Variables

```{r}
# Require observed values for all three specified variables.
required_vars <- c("occupation", "workclass", "native_country")

adult_complete <- adult_hours[
  complete.cases(adult_hours[required_vars]),
]

data.frame(
  Before = nrow(adult_hours),
  Removed = nrow(adult_hours) - nrow(adult_complete),
  After = nrow(adult_complete)
)
```

This step removes `r nrow(adult_hours) - nrow(adult_complete)`
observations and leaves `r nrow(adult_complete)` observations.
The removed records lack at least one of the three variables required
for the analysis. Restricting the sample to complete records allows
these variables to be used consistently. However, this reduces the
sample size and may introduce selection bias if missingness is
systematically related to individual characteristics.

## (f) Apply Age and Education Restrictions

```{r}
# Retain ages 18 through 90, inclusive, and exclude Preschool.
adult_clean <- subset(
  adult_complete,
  age >= 18 & age <= 90 & education != "Preschool"
)

data.frame(
  Before = nrow(adult_complete),
  Removed = nrow(adult_complete) - nrow(adult_clean),
  After = nrow(adult_clean)
)
```

Individuals younger than 18, older than 90, or whose education is
recorded as Preschool are excluded. Ages 18 and 90 are retained.
This step removes `r nrow(adult_complete) - nrow(adult_clean)`
observations.

## (g) Final Sample Size

```{r}
# Summarize the sample size after each cleaning step.
sample_flow <- data.frame(
  Stage = c(
    "Original training data",
    "Working at least 10 hours per week",
    "Complete occupation, workclass, and native country",
    "Ages 18-90 and education other than Preschool"
  ),
  Observations = c(
    nrow(adult),
    nrow(adult_hours),
    nrow(adult_complete),
    nrow(adult_clean)
  )
)
knitr::kable(sample_flow, row.names = FALSE)

# Verify that the final sample satisfies all required conditions.
stopifnot(
  all(adult_clean$hours_per_week >= 10),
  all(complete.cases(adult_clean[required_vars])),
  all(adult_clean$age >= 18 & adult_clean$age <= 90),
  all(adult_clean$education != "Preschool")
)
```

The final cleaned sample contains **`r nrow(adult_clean)` observations**.


# Problem 3: Happy Numbers

## (a) Compute the Next Value

```{r}
#' Compute the next value in a happy-number sequence.
#'
#' @param n A single positive integer no greater than
#'   .Machine$integer.max.
#' @return The sum of the squared decimal digits of n.
#' @details Invalid inputs produce an informative error.
nextHappy <- function(n) {
  if (!is.numeric(n) || is.complex(n) || length(n) != 1L) {
    stop("n must be a single positive integer.")
  }

  if (is.na(n) || !is.finite(n) || n <= 0 || n != floor(n)) {
    stop("n must be a single positive integer.")
  }

  if (n > .Machine$integer.max) {
    stop("n must not exceed .Machine$integer.max.")
  }

  digits <- as.integer(strsplit(as.character(as.integer(n)), "")[[1]])
  sum(digits^2)
}

# Reproduce the examples.
nextHappy(19)
nextHappy(82)
```

The function splits the input into decimal digits, squares each digit,
and adds the results. For example, $19$ gives $1^2 + 9^2 = 82$,
and $82$ gives $8^2 + 2^2 = 68$.
The supported input range is 1 through `r .Machine$integer.max`.

```{r}
# Demonstrate informative errors without interrupting rendering.
tryCatch(nextHappy(0), error = conditionMessage)
tryCatch(nextHappy(2.5), error = conditionMessage)
tryCatch(nextHappy("19"), error = conditionMessage)
```

## (b) Generate the Sequence and Detect Happiness

```{r}
#' Generate a happy-number sequence and determine whether it is happy.
#'
#' @param n A single positive integer in the range supported by nextHappy().
#' @return A list containing sequence, a numeric vector, and isHappy,
#'   a logical value indicating whether the sequence reaches 1.
#' @details The sequence includes the starting value and ends at 1
#'   for happy numbers. For unhappy numbers, it stops before appending
#'   the first repeated value. Input validation is performed by nextHappy().
happySequence <- function(n) {
  # Validate the input using nextHappy().
  nextHappy(n)

  sequence <- n

  repeat {
    current <- sequence[length(sequence)]

    if (current == 1) {
      return(list(sequence = sequence, isHappy = TRUE))
    }

    next_value <- nextHappy(current)

    if (next_value %in% sequence) {
      return(list(sequence = sequence, isHappy = FALSE))
    }

    sequence <- c(sequence, next_value)
  }
}

# Reproduce the examples.
happySequence(19)
happySequence(4)

# Check the case where the starting value is already 1.
happySequence(1)

# Demonstrate input validation.
tryCatch(happySequence(-4), error = conditionMessage)
```

Starting from 19 produces the sequence
$19 \rightarrow 82 \rightarrow 68 \rightarrow 100 \rightarrow 1$,
so 19 is happy.

Starting from 4 produces
$4 \rightarrow 16 \rightarrow 37 \rightarrow 58 \rightarrow 89
\rightarrow 145 \rightarrow 42 \rightarrow 20$.
The next value would be 4 again, so the sequence cycles and 4 is unhappy.
The repeated value is not appended to the returned sequence.

## (c) Analyze Integers from 1 Through 1000

Sequence length is defined as the number of values returned by
`happySequence()`, including the starting value and the terminal 1
for happy numbers. For unhappy numbers, the first repeated value
is not counted again. Both happy and unhappy starting values are
included when finding the longest sequence.

```{r}
# Generate the sequence for every starting value.
starting_values <- 1:1000
results <- lapply(starting_values, happySequence)

# Extract happiness indicators and sequence lengths.
is_happy <- logical(length(starting_values))
sequence_lengths <- integer(length(starting_values))

for (i in seq_along(starting_values)) {
  is_happy[i] <- results[[i]]$isHappy
  sequence_lengths[i] <- length(results[[i]]$sequence)
}

# Count happy numbers and calculate their percentage.
happy_count <- sum(is_happy)
happy_percentage <- 100 * mean(is_happy)

data.frame(
  Happy_Count = happy_count,
  Happy_Percentage = happy_percentage
)
```

Among the integers from 1 through 1000, **`r happy_count` are happy**,
representing **`r happy_percentage`%** of the integers considered.

```{r}
# Identify all starting values tied for the longest sequence.
max_length <- max(sequence_lengths)
longest_starts <- starting_values[sequence_lengths == max_length]

longest_table <- data.frame(
  Starting_Value = longest_starts,
  Sequence_Length = sequence_lengths[sequence_lengths == max_length],
  Is_Happy = is_happy[sequence_lengths == max_length]
)
knitr::kable(longest_table, row.names = FALSE)

# Display the sequence for the smallest starting value among the ties.
happySequence(min(longest_starts))
```

The longest sequence contains **`r max_length` values** under the
counting convention stated above. The starting values attaining this
length are **`r paste(longest_starts, collapse = ", ")`**.
The output above shows the sequence for the smallest of these starting
values, `r min(longest_starts)`.

# AI Assistance Disclosure

OpenAI Codex (September 22, 2026) was used during the preparation of
this assignment. It provided R code and draft written
responses for Problems 3, as well as guidance on Quarto debugging. I incorporated this material into the current draft and ran the code in RStudio to inspect the outputs.
